---
title: "Fitting parametric survival models in R"
author: "Roche"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
csl: biomedicine.csl
vignette: >
  %\VignetteIndexEntryFitting parametric survival models in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  message = FALSE
)
```

<style type="text/css">
  
  body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
    font-size: 10px;
}
h1.title {
  font-size: 38px;
}
h1 { /* Header 1 */
    font-size: 28px;
}
h2 { /* Header 2 */
    font-size: 22px;
}
h3 { /* Header 3 */
    font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>
  
# Introduction
Parametric survival models are often the preferred method of extrapolating
survival data for use in economic models. The National Institute for Health and
Care Excellence (NICE) Decision Support Unit (DSU) technical support document
(TSD) 14 recommends that the Exponential, Weibull, Gompertz, log-logistic, log
normal and Generalized Gamma parametric models should all be
considered.[@latimer2011nice] More recently, NICE also discusses more flexible
models in NICE DSU TSD 21, however, more these models are not in the scope of
this package.[@rutherford2020nice] The Canadian Agency for Drugs and
Technologies in Health (CADTH) additionally specifies that the Gamma
distribution must also be considered. This document therefore details the
characteristics of each of these distributions and demonstrates how the
parameters from each distribution, outputted using the flexsurvPlus package, can
be implemented within an economic model.[@cadth2020] The Generalized F
distribution is not commonly used, however it has been included in this package
in case it is required.

The flexsurvPlus package allows the inclusion of a treatment effect in the
following three ways:
  
  * Separate models - Models fitted to each treatment arm separately

*  Independent shape models - Models fitted to both/all treatment arms including
a treatment covariate to model the effect of treatment on both the scale and
shape parameter(s) of the distribution.

*  Common shape models - Models fitted to both/all treatment arms including a
treatment covariate to model the effect of treatment on the scale parameter of
the distribution. The shape parameter(s) of the distribution is common across
treatments which reflects an assumption of proportional hazards or accelerated
failure time between treatments depending on the distribution

This document details how to use the <tt>flexsurvPlus</tt> package to perform
these models. A separate vignette; "Parametric survival analysis using the
flexsurvPlus package: understanding the theory" details the theory behind the
models.


# Set up packages and data

## Install packages

The following packages are required to run this example:
  
```{r setup}
# Libraries
library(flexsurvPlus)
library(tibble)
library(dplyr)
library(survival)
library(survminer)
library(tidyr)
library(boot)

# Non-scientific notation
options(scipen=999) 

# Colours for plots
blue = rgb(69, 185, 209, max=255)
pink = rgb(211,78,147,max=255)
Dyellow = rgb(214, 200, 16, max=255)
orange<-rgb(247,139,21,max=255)

```

## Read in the data

To perform survival analyses, patient level data is required for the survival
endpoints.

This example uses a standard simulated data set (adtte). There is no standard
naming that is needed for this package however, there are some set variables
that are needed:
  
* Time - a numeric variable
* Event - a binary variable (event=1, censor=0)
* Treatment - a character variable with the name of the intervention treatment 

The data must be in "wide" format such that there is one row per patient and
columns for each endpoint separately. In this example, we analyze
progression-free survival (PFS).
```{r}
adtte <- sim_adtte(seed = 2020, rho = 0.6)
head(adtte)

# subset PFS data and rename
PFS_data <- adtte %>%
  filter(PARAMCD=="PFS") %>%
  transmute(USUBJID,
            ARMCD,
            PFS_days = AVAL,
            PFS_event = 1- CNSR
  )
head(PFS_data)

```

# Exploratory analysis
Before performing any statistical analysis, it is important to explore the data.
Most importantly is a Kaplan-Meier plot. A simple Kaplan-Meier plot has been
produced here:


```{r, warning=FALSE}

# Create survfit object
km.est.PFS <- survfit(Surv(PFS_days, PFS_event) ~ ARMCD, 
                      data = PFS_data, 
                      conf.type = 'plain')

# Plot Kaplan-Meier
plot(km.est.PFS, 
     col = c(blue, pink), # plot colours
     lty = c(1:2), # line type
     xlab = "Time (Days)", # x-axis label
     ylab = "Progression-free survival", # y-axis label
     xlim = c(0, 800)) 

legend(x = 500, 
       y = .9, 
       legend = c("Arm A", "Arm B"), 
       lty = c(1:2), 
       col = c(blue, pink))

```

# Fitting the models

The <tt>runPSM</tt> function fits parametric survival models for multiple
distributions using the <tt>flexsurv</tt> package (using the
<tt>flexsurvPlus</tt> function <tt>fit_models</tt>),  manipulates the flexsurv
objects to get the parameter estimates and AIC and BIC value (using the
<tt>flexsurvPlus</tt> function <tt>get_params</tt>) and rearranges the parameter
estimates such that they can easily be output to excel to calculate survival for
both the intervention and reference treatment in an economic model.

These functions can be used to estimate 3 types of model:
  
*  Separate models - Models fitted to each treatment arm separately
    *  These models are fitted using the <tt>runPSM</tt> function using the
    <tt>model.type="Separate"</tt> argument.

*  Independent shape models - Models fitted to both/all treatment arms including
a treatment covariate to model the effect of treatment on both the scale and
shape parameter(s) of the distribution.
    * They have been fit using the <tt>runPSM</tt> function using the 
    <tt>model.type="Independent shape"</tt> argument.

*  Common shape models - Models fitted to both/all treatment arms including a
treatment covariate to model the effect of treatment on the scale parameter of
the distribution. The shape parameter(s) of the distribution is common across
treatments which reflects an assumption of proportional hazards or accelerated
failure time between treatments depending on the distribution
    * These models have been fit using the <tt>runPSM</tt> function using  the
    <tt>model.type="Common shape"</tt> argument.


The inputs to the <tt>runPSM</tt> function are:

*  <tt>data</tt> - A data frame containing individual patient data for the
relevant time to event outcomes

*  <tt>time_var</tt> - Name of time variable in 'data'. Variable must be numerical
and >0.

*  <tt>event_var</tt> - Name of event variable in 'data'. Variable must be
numerical and contain 1's to indicate an event and 0 to indicate a censor.

*  <tt>model.type</tt> - Character vector indicating the types of model formula
provided. Permitted values are 'Common shape', 'Independent shape' or 'Separate'
as per the models explained above.

* <tt>distr</tt> - Each type of model can be fitted with multiple distributions.
The distributions available for this package are:

    * Exponential ('exp')

    * Weibull ('weibull')

    * Gompertz ('gompertz')

    * Log-normal ('lnorm')

    * Log-logistic ('llogis')

    * Generalized gamma ('gengamma')

    * Gamma ('gamma')

    * Generalized F ('genf')

*  <tt>strata_var</tt> - Name of stratification variable in "data". This is
usually the treatment variable and must be categorical.

*  <tt>int_name</tt> - Character to indicate the name of the treatment of
interest, must be a level of the "strata_var" column in "data", used for
labelling the parameters.

*  <tt>ref_name </tt> - Character to indicate the name of the reference treatment,
must be a level of the "strata_var" column in "data", used for labelling the
parameters.

More information about each function can be used by running the code <tt>?runPSM</tt>.

Example code to fit all models is presented below.


```{r}
psm_PFS_all <- runPSM(data=PFS_data,
                     time_var="PFS_days",
                     event_var="PFS_event",
                     model.type= c("Common shape", 
                                   "Independent shape", 
                                   "Separate"),
                     distr = c('exp',
                               'weibull',
                               'gompertz',
                               'lnorm',
                               'llogis',
                               'gengamma',
                               'gamma',
                               'genf'),
                     strata_var = "ARMCD",
                     int_name="A",
                     ref_name = "B")
psm_PFS_all

```




# Estimate survival from the models and plot the curves

Survival at a given time can be estimated using the .

This can be used to cross check survival in excel models. The example is given
below selects only a few models for clarity.


```{r}
# Landmark survival

# vector of times to estimate survival (days)
time = c(0, 100, 200, 300)

# Example 1: intervention arm, weibull distribution, common shape model
1 - pweibull(time,
         shape = psm_PFS_all$parameters[psm_PFS_all$parameters$Model=="Common shape",]$weibull.shape.int,
         scale = psm_PFS_all$parameters[psm_PFS_all$parameters$Model=="Common shape",]$weibull.scale.int)



```



The <tt>get_curvefits</tt> function can also be used to generate the data
required to plot the survival curves, overlaid on top of the KM plot. The time
argument should reflect how long you want to the extrapolate for and the unit of
time is the same as the input data (in this example, days).

The <tt>ggsurvplot</tt> function from the <tt>survminer</tt> package has been
used to show how the survival curves can be plotted.

```{r}

# for clarity just pick the 2 different weibull models and a Gamma. 
selected_models <- c("comshp.weibull","comshp.gengamma","indshp.gengamma")

# Create data set for plots
curvefits_PFS <- get_curvefits(models = psm_PFS_all$models[selected_models],
                                  time = seq(from=0, to = 500, by = 10))

# format data ready for the plot
curvefits_df_PFS <- bind_rows(curvefits_PFS$curvefits, .id = "Dist") %>%
  pivot_wider(names_from  = c(ARM), values_from  = c(est))

analysis_data_TrtA <- filter(PFS_data, ARMCD=="A")
analysis_data_TrtB <- filter(PFS_data, ARMCD=="B")

# Get KM estimates
km.est.PFS.TrtA <- survfit(Surv(PFS_days, PFS_event) ~ 1 , data = analysis_data_TrtA, conf.type = 'plain')
km.est.PFS.TrtB <- survfit(Surv(PFS_days, PFS_event) ~ 1 , data = analysis_data_TrtB, conf.type = 'plain')


# Plot survival curves for arm A
plot.PFS.A <- ggsurvplot(km.est.PFS.TrtA,
                        combine = TRUE,
                        censor = FALSE,
                        risk.table = TRUE,
                        conf.int = FALSE,
                        break.x.by = 100,
                        xlim = c(0, 500),
                        xlab = "Time (days)",
                        size = 0.72,
                        linetype = c(6,6,6,6, 1, 6, 6,6,6),
                        title  =  'Kaplan-Meier of OS with standard distribution overlays',
                        legend.title = '',
                        legend.labs = c("KM"),
                        surv.median.line = 'hv',
                        palette = c(rgb(0,0,0,max=30)),
                        risk.table.y.text.col = F
)


plot.PFS.A$plot <- plot.PFS.A$plot +
  geom_line(aes(x = time, y = Int,
                colour = Dist,
                linetype = Dist),
            size = 0.72,
            data = curvefits_df_PFS) +
  scale_color_manual(values = c(blue, pink, Dyellow, orange))

plot.PFS.A

# Plot survival curves for arm B
plot.PFS.B <- ggsurvplot(km.est.PFS.TrtB,
                        combine = TRUE,
                        censor = FALSE,
                        risk.table = TRUE,
                        conf.int = FALSE,
                        break.x.by = 100,
                        xlim = c(0, 500),
                        xlab = "Time (days)",
                        size = 0.72,
                        linetype = c(6,6,6,6, 1, 6, 6,6,6),
                        title  =  'Kaplan-Meier of OS with standard distribution overlays',
                        legend.title = '',
                        legend.labs = c("KM"),
                        surv.median.line = 'hv',
                        palette = c(rgb(0,0,0,max=30)),
                        risk.table.y.text.col = F
)


plot.PFS.B$plot <- plot.PFS.B$plot +
  geom_line(aes(x = time, y = Ref,
                colour = Dist,
                linetype = Dist),
            size = 0.72,
            data = curvefits_df_PFS) +
  scale_color_manual(values = c(blue, pink, Dyellow, orange))

plot.PFS.B

```


# Adressing uncertainty - Bootstrapping

Bootstrapping has been used to estimate the uncertainty of the parameters from the survival models. Boostrapping is used for two reasons motivated by intent of this package to support further modeling in excel.
1. To simplify and accelerate calculations in excel while maintaining correlations between parameters (as is commonly done for NMA)
2. To maintain correlations across multiple endpoints (see separate vignette for details)

Bootstrapping involves:

1. Sampling, with replacement, from all patients
2. Estimating all specified parametric survival models

This procedure is repeated multiple times to obtain a distribution of parameters.
For this example, bootstrap estimates of the parameters were calculated using the
<tt>boot</tt> package. An argument for the <tt>boot</tt> function is
<tt>statistic</tt> which is a function which when applied to data returns a
vector containing the statistic(s) of interest. The <tt>bootPSM</tt> function 
in the <tt>flexsurvPlus</tt> package can be used for this purpose.

The inputs for the <tt>bootPSM</tt> function are identical to the <tt>runPSM</tt> function, however there is one additional argument:
*<tt>i</tt> - Index used to select a sample within <tt>boot</tt>


```{r}
set.seed(2358)
# for speed and illustration only 4 models are selected for bootstrapping - all could be included
n.sim <- 100
PSM_bootstraps_PFS <- boot(
  statistic = bootPSM, # bootstrap function
  R=n.sim, # number of bootstrap samples
  data=PFS_data,
  time_var="PFS_days",
  event_var="PFS_event",
  model.type = c("Common shape", "Separate"),
  distr = c('weibull', 'gamma'),
  strata_var = "ARMCD",
  int_name = "B",
  ref_name = "A"
)

```

To simplify the post processing of the bootstrap samples a function \code{\link{bootPSMtidy}} is provided that does some manipulations. This returns a tibble similar to the \code{Parameters} one that is returned by \code{\link{bootPSMtidy}}. SampleID identified the bootstrap sample. SampleID of 0 are the parameters for the original data.


```{r}
PSM_bootstrap_PFS_data <- bootPSMtidy(PSM_bootstraps_PFS)
head(PSM_bootstrap_PFS_data)

```

# Estimating quantities from the sample

As the <tt>flexsurv</tt> parameterisations are used any quantity of interest can be simply calculated for all models and samples through use of the flexsurv functions such as the extrapolated means via flexsurv::mean_weibull.

```{r}
estimated_deltamean <- PSM_bootstrap_PFS_data %>%
  transmute(SampleID, Model,
            Weibull.Mean.ref = flexsurv::mean_weibull(shape = weibull.shape.ref, scale = weibull.scale.ref),
            Weibull.Mean.int = flexsurv::mean_weibull(shape = weibull.shape.int, scale = weibull.scale.int),
            Weibull.Mean.delta = Weibull.Mean.int - Weibull.Mean.ref)

head(estimated_deltamean)

estimated_deltamean %>%
  ggplot(aes(x = Weibull.Mean.delta)) + 
  geom_density()

```



# Outputing parameters to excel 

The primary use of the bootstrap samples is to be used in probabilistic sensitivity analyses in economic models. 

Once all the models have been fit and bootstrap samples estimated they can be output to excel. By selecting SampleID = 0 the estimates for the original data are returned. To run PSA in excel only a random number between 1 and the number of samples needs to be generated.

```{r}

# Add meta data to include details of analysis
parameters_PFS <- PSM_bootstrap_PFS_data %>%
  mutate(Study_name = "Study ABC",
         Datacut = "Final",
         Population = "ITT",
         Endpoint = "PFS")

#write.csv(parameters_PFS, "params_for_model.csv")

```


# References
