---
title: "Fitting parametric survival models in R"
author: "Roche"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
csl: biomedicine.csl
vignette: >
  %\VignetteIndexEntryFitting parametric survival models in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 10px;
}
h1.title {
  font-size: 38px;
}
h1 { /* Header 1 */
  font-size: 28px;
  }
h2 { /* Header 2 */
    font-size: 22px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# Introduction
Parametric survival models are often the preferred method of extrapolating
survival data for use in economic models. The National Institute for Health and
Care Excellence (NICE) Decision Support Unit (DSU) technical support document
(TSD) 14 recommends that the Exponential, Weibull, Gompertz, log-logistic, log
normal and Generalized Gamma parametric models should all be
considered.[@latimer2011nice] More recently, NICE also discusses more flexible
models in NICE DSU TSD 21, however, more these models are not in the scope of
this package.[@rutherford2020nice] The Canadian Agency for Drugs and
Technologies in Health (CADTH) additionally specifies that the Gamma
distribution must also be considered. This document therefore details the
characteristics of each of these distributions and demonstrates how the
parameters from each distribution, outputted using the flexsurvPlus package, can be
implemented within an economic model.[@cadth2020]

The flexsurvPlus package allows the inclusion of a treatment effect in the following
three ways:

* Separate models - Models fitted to each treatment arm separately
  
*  Independent shape models - Models fitted to both/all treatment arms including
a treatment covariate to model the effect of treatment on both the scale and
shape parameter(s) of the distribution.
    
*  Common shape models - Models fitted to both/all treatment arms including a
treatment covariate to model the effect of treatment on the scale parameter of
the distribution. The shape parameter(s) of the distribution is common across
treatments which reflects an assumption of proportional hazards or accelerated
failure time between treatments depending on the distribution
    
This document details how to use the <tt>flexsurvPlus</tt> package to perform these models. A separate vignette; "Parametric survival analysis using the flexsurvPlus package: understanding the theory" details the theory behind the models.


# Set up packages and data

## Install packages

The following packages are required to run this example:

```{r setup}
library(flexsurvPlus)
library(tibble)
library(dplyr)
library(survival)
library(survminer)
library(tidyr)
library(boot)


blue = rgb(69, 185, 209, max=255)
red = rgb(225, 55, 60, max=255)
yellow = rgb(238, 224, 30, max=255)
pink = rgb(211,78,147,max=255)
Dblue = rgb(0,45,92,max=255)
Dyellow = rgb(214, 200, 16, max=255)
green = rgb(0,1,0)
Dyellow = rgb(214, 200, 16, max=255)
black<-rgb(0,0,0,max=255)
grey1<-rgb(0,0,0,max=30)
grey2<-rgb(10,10,10,max=30)
grey3<-rgb(20,20,20,max=30)
grey3_dup<-rgb(19,20,20,max=30)
lightPink<-rgb(229,151,192,max=255)
orange<-rgb(247,139,21,max=255)

```

## Read in the data

To perform survival analyses, patient level data is required for the survival endpoints. 

This example uses a standard simulated data set
(adtte). There is no standard naming that is
needed for this package however, there are some set variables that are needed:

* Time - a numeric variable
* Event - a binary variable (event=1, censor=0)
* Treatment - a character variable with the name of the intervention treatment 

The data must be in "wide" format such that there is one row per patient and columns for each endpoint separately.
In this example, we use overall survival (OS) and progression-free survival (PFS). 
```{r}


adtte <- sim_adtte(seed = 2020, rho = 0.6)

# subset OS data and rename
OS_data <- adtte %>%
  filter(PARAMCD=="OS") %>%
  transmute(USUBJID,
            ARMCD,
            OS_days = AVAL,
            OS_event = 1- CNSR
  )

# subset PFS data and rename
PFS_data <- adtte %>%
  filter(PARAMCD=="PFS") %>%
  transmute(USUBJID,
            ARMCD,
            PFS_days = AVAL,
            PFS_event = 1- CNSR
  )

analysis_data <- left_join(OS_data, PFS_data, by = c("USUBJID", "ARMCD"))


```

# Exploratory analysis
Before performing any statistical analysis, it is important to explore the data.

Most importantly is a Kaplan-Meier plot and code to do this using <tt>ggsurvplot</tt> is presented below.


```{r, warning=FALSE}

km.est.OS <- survfit(Surv(OS_days, OS_event) ~ ARMCD , data = analysis_data, conf.type = 'plain')
km.est.PFS <- survfit(Surv(PFS_days, PFS_event) ~ ARMCD , data = analysis_data, conf.type = 'plain')

#KM_list <- list(OS = km.est.OS, PFS = km.est.PFS)

KM_plot_OS <- ggsurvplot(km.est.OS, risk.table = TRUE, data = analysis_data,
           break.time.by = 200,
           conf.int = FALSE,
           censor=FALSE,
           legend.title = '',
           xlab = paste0('Overall survival (Days)'),
           size = 0.72,
           xlim = c(0, 800))
KM_plot_OS

KM_plot_PFS <- ggsurvplot(km.est.PFS, risk.table = TRUE, data = analysis_data,
           break.time.by = 200,
           conf.int = FALSE,
           censor=FALSE,
           legend.title = '',
           xlab = paste0('Progression-free survival (Days)'),
           size = 0.72,
           xlim = c(0, 800))
KM_plot_PFS

```

# Fitting the models

The <tt>runPSM</tt> function fits parametric survival models for multiple distributions using the <tt>flexsurv</tt> package (using the <tt>flexsurvPlus</tt> function <tt>fit_models</tt>),  manipulates the flexsurv objects to get the parameter estimates and AIC and BIC value (using the <tt>flexsurvPlus</tt> function <tt>get_params</tt>) and rearranges the parameter estimates such that they can easily be output to excel to calculate survival for both the intervention and reference treatment in an economic model.



These functions can be used to estimate 3 types of model:

* Separate models - Models fitted to each treatment arm separately
  
*  Independent shape models - Models fitted to both/all treatment arms including
a treatment covariate to model the effect of treatment on both the scale and
shape parameter(s) of the distribution.
    
*  Common shape models - Models fitted to both/all treatment arms including a
treatment covariate to model the effect of treatment on the scale parameter of
the distribution. The shape parameter(s) of the distribution is common across
treatments which reflects an assumption of proportional hazards or accelerated
failure time between treatments depending on the distribution


Each type of model can be fitted with a multiple distributions. The distributions available for this package are:
* Exponential ('exp')
* Weibull ('weibull')
* Gompertz ('gompertz')
* Log-normal ('lnorm')
* Log-logistic ('llogis')
* Generalized gamma ('gengamma')
* Gamma ('gamma')


For more information on the different types of models, please see the survival analysis theory vignette included in the <tt>flexsurvPlus</tt> package.

The inputs to the <tt>runPSM</tt> function are:
*<tt>data</tt> - A data frame containing individual patient data for the relevant time to event outcomes
*<tt>time_var</tt> - Name of time variable in 'data'. Variable must be numerical and >0.
*<tt>event_var</tt> - Name of event variable in 'data'. Variable must be numerical and contain 1's to indicate an event and 0 to indicate a censor.
*<tt>model.type</tt> - Character vector indicating the types of model formula provided. Permitted values are 'Common shape', 'Independent shape' or 'Separate' as per the models explained above.
*<tt>strata_var</tt> - Name of stratification variable in "data". This is usually the treatment variable and must be categorical.
*<tt>int_name</tt> - Character to indicate the name of the treatment of interest, must be a level of the "strata_var" column in "data", used for labelling the parameters.
*<tt>ref_name </tt> - Character to indicate the name of the reference treatment, must be a level of the "strata_var" column in "data", used for labelling the parameters.

More information about each function can be used by running the code ?runPSM.

Example code to fit all models is presented below.


```{r}

psm_OS_all <- runPSM(data=analysis_data,
                     time_var="OS_days",
                     event_var="OS_event",
                     model.type= c("Common shape", 
                                   "Independent shape", 
                                   "Separate"),
                     distr = c('exp',
                               'weibull',
                               'gompertz',
                               'lnorm',
                               'llogis',
                               'gengamma',
                               'gamma'),
                     strata_var = "ARMCD",
                     int_name="A",
                     ref_name = "B")
psm_OS_all

psm_PFS_all <- runPSM(data=analysis_data,
                     time_var="PFS_days",
                     event_var="PFS_event",
                     model.type= c("Common shape", 
                                   "Independent shape", 
                                   "Separate"),
                     distr = c('exp',
                               'weibull',
                               'gompertz',
                               'lnorm',
                               'llogis',
                               'gengamma',
                               'gamma'),
                     strata_var = "ARMCD",
                     int_name="A",
                     ref_name = "B")
psm_PFS_all


```


## Separate models for each treatment

These models have been fit using the <tt>runPSM</tt> function using the model.type="Separate" argument.

## Independent shape models 

These models are similar to the separate model for each treatment as they allow all parameters to vary by treatment, however it is performed using 1 model for 2 treatments, rather than 2 models as above. They have been fit using the <tt>runPSM</tt> function using the model.type="Independent shape" argument.


## Fit common shape models
These models have been fit using the <tt>runPSM</tt> function using  the model.type="Common shape" argument.

## Output parameters to excel

Once all the models have been fit, they can be output to excel.

```{r}
# Add meta data to add in details of analysis
parameters_OS <- psm_OS_all$parameters %>%
  mutate(Study_name = "Study ABC",
         Datacut = "Final",
         Population = "ITT",
         Endpoint = "OS")

# Add meta data to add in details of analysis
parameters_PFS <- psm_PFS_all$parameters %>%
  mutate(Study_name = "Study ABC",
         Datacut = "Final",
         Population = "ITT",
         Endpoint = "PFS")

# stack the PFS & OS parameters
parameters_all <- rbind(parameters_PFS, parameters_OS)

parameters_all

#write.csv(parameters_all, "params_for_model.csv")

```

# Estimate survival from the models and plot the curves

Survival at a given time can be estimated using the <tt>get_curvefits</tt> function as shown below for treatment-effect models. 

This can be used to cross check survival in excel models.  The example is given below selects only a few models for clarity.


```{r}
# Landmark survival

# for clarity just pick the 4 different weibull models. 
selected_models <- c("weibull.comshp","weibull.indshp","weibull.sepint", "weibull.sepref")

curvefits_OS_landmark <- get_curvefits(models = psm_OS_all$models[selected_models],
                                           time = c(0, 100, 200, 300))
curvefits_df_OS_landmark <- bind_rows(curvefits_OS_landmark$curvefits, .id = "Dist")  %>%
  pivot_wider(names_from  = c(Dist, ARM), values_from  = est) %>%
  as.data.frame()
curvefits_df_OS_landmark
```



The <tt>get_curvefits</tt> function can also be used to generate the data required to plot the survival curves, overlaid on top of the KM plot.
The time argument should reflect how long you want to the extrapolate for and the unit of time is the same as the input data (in this example, days).

 The <tt>ggsurvplot</tt> function from the <tt>survminer</tt> package has been used to show how the survival curves can be plotted.

```{r}

# for clarity just pick the 2 different weibull models and a Gamma. 
selected_models <- c("weibull.comshp","weibull.indshp","gamma.comshp")

# Create data set for plots
curvefits_PFS <- get_curvefits(models = psm_PFS_all$models[selected_models],
                                  time = seq(from=0, to = 500, by = 10))

# format data ready for the plot
curvefits_df_PFS <- bind_rows(curvefits_PFS$curvefits, .id = "Dist") %>%
  pivot_wider(names_from  = c(ARM), values_from  = c(est))

analysis_data_TrtA <- filter(analysis_data, ARMCD=="A")
analysis_data_TrtB <- filter(analysis_data, ARMCD=="B")

# Get KM estimates
km.est.PFS.TrtA <- survfit(Surv(PFS_days, PFS_event) ~ 1 , data = analysis_data_TrtA, conf.type = 'plain')
km.est.PFS.TrtB <- survfit(Surv(PFS_days, PFS_event) ~ 1 , data = analysis_data_TrtB, conf.type = 'plain')


# Plot survival curves for arm A
plot.PFS.A <- ggsurvplot(km.est.PFS.TrtA,
                        combine = TRUE,
                        censor = FALSE,
                        risk.table = TRUE,
                        conf.int = FALSE,
                        break.x.by = 100,
                        xlim = c(0, 500),
                        xlab = "Time (days)",
                        size = 0.72,
                        linetype = c(6,6,6,6, 1, 6, 6,6,6),
                        title  =  'Kaplan-Meier of OS with standard distribution overlays',
                        legend.title = '',
                        legend.labs = c("KM"),
                        surv.median.line = 'hv',
                        palette = c(rgb(0,0,0,max=30)),
                        risk.table.y.text.col = F
)


plot.PFS.A$plot <- plot.PFS.A$plot +
  geom_line(aes(x = time, y = Int,
                colour = Dist,
                linetype = Dist),
            size = 0.72,
            data = curvefits_df_PFS) +
  scale_color_manual(values = c(blue, pink, Dyellow, orange))

plot.PFS.A

# Plot survival curves for arm B
plot.PFS.B <- ggsurvplot(km.est.PFS.TrtB,
                        combine = TRUE,
                        censor = FALSE,
                        risk.table = TRUE,
                        conf.int = FALSE,
                        break.x.by = 100,
                        xlim = c(0, 500),
                        xlab = "Time (days)",
                        size = 0.72,
                        linetype = c(6,6,6,6, 1, 6, 6,6,6),
                        title  =  'Kaplan-Meier of OS with standard distribution overlays',
                        legend.title = '',
                        legend.labs = c("KM"),
                        surv.median.line = 'hv',
                        palette = c(rgb(0,0,0,max=30)),
                        risk.table.y.text.col = F
)


plot.PFS.B$plot <- plot.PFS.B$plot +
  geom_line(aes(x = time, y = Ref,
                colour = Dist,
                linetype = Dist),
            size = 0.72,
            data = curvefits_df_PFS) +
  scale_color_manual(values = c(blue, pink, Dyellow, orange))

plot.OS.B

```


#References
